A spatial analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-19 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Continued editing section on selection of variables.

  \begin{abstract}
  Spreading with astonishing speed, the novel SARS-CoV2 has swept the
  globe, causing enormous stress to health systems and prompting social
  distance guidelines and mandates to arrest its progress. While there is
  encouraging evidence that early public health interventions have slowed
  the spread of the virus, this has come at a high cost as the global
  economy is brought to its knees. How and when to ease restrictions to
  movement hinges in part on the question whether SARS-CoV2 will display
  seasonality associated with variations in temperature, humidity, and
  hours of sunshine. In this research, we address this question by means
  of a spatial analysis of the incidence of COVID-19 in the provinces in
  Spain. Use of a spatial Seemingly Unrelated Regressions (SUR) approach
  allows us to model the incidence of reported cases of the disease per
  100,000 population, as a function of temperature and humidity, while
  controlling for GDP per capita, population density, percentage of older
  adults in the population, and presence of mass transit systems. An
  interesting aspect of the spatial SUR approach is that it models
  incidence as a contagion process. Our results indicate that incidence of
  the disease is lower at higher temperatures and higher levels of
  humidity, although coefficients for this variable are significant only
  in some equations. Sunshine, in contrast, displays a positive
  association with incidence of the disease. Our control variables also
  yield interesting insights. Higher incidence is associated with higher
  GDP per capita and presence of mass transit systems in the province; in
  contrast, population density and percentage of older adults display
  negative associations with incidence of COVID-19.\\
  \end{abstract}

@@

Some of what is thought about the possible seasonality of COVID-19 is
based on analogies to the patterns of other known respiratory viruses.
However, de Ángel Solá et al.~(2020) note that ``not all seasonal
respiratory viruses experience the same spatiotemporal patterns''
(section 4). This urges caution when extrapolating from known viruses,
and indeed, the evidence in this respect is inconclusive. At a global
scale, de Ángel Solá et al.~(2020) see less risk in the Caribean Basin;
however, Coelho et al.~(Coelho et al., 2020) warn that at least during
the exponential phase, expansion of the virus is not driven by climate.
Similarly, whereas Araujo and Naimi (2020) argue that spread of
SARS-CoV2 will likely be constrained by climate, Harbert et al.~(2020)
remain unconvinced that spatial modelling can currently discriminate the
distribution of the disease on the basis of climate, at least in the
United States. Yao et al.~({\textbf{???}}), examined data from China and
came to the conclusion that neither temperature nor ultraviolet indices
had an association with transmission of COVID-19. This is despite
previous research that has linked less exposure to UVB radiation to
higher prevalence and severity of acute respiratory tract infections
(Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen et
al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
Moriyama, Hugentobler, and Iwasaki 2020). To further complicate matters,
much of the relevant work has yet to be peer-reviewed and therefore is
open to change (see for example the challenge of Harbert et al. (2020)
to Araujo and Naimi (2020)). In the United States, the National Academy
of Sciences, Engineering, and Medicine was engaged by the Office of the
Executive for guidance on this matter (see National Academies of
Sciences, Engineering and Medicine, 2020). Their conclusion summarizes
the situation well (see p.~6): ``Some limited data support a potential
waning of cases in warmer and more humid seasons, yet none are without
major limitations\ldots{}Additional studies as the SARS-CoV-2 pandemic
unfolds could shed more light on the effects of climate on
transmission.''

@@

\begin{quote}
https://github.com/paezha/covid19-environmental-correlates
\end{quote}

@@

Two set of variables has been selected to explain the diary incidence
ratio in each Spanish province. The first, are climatic variables
(temperature, humidity and sunshine) and its impact has been argued in
the introduction section. The second, is a set of specific
characteristics of each province. The objective is include in our model
some control variables (socioeconomic characteristics) to capture the
probable heterogeneity of each spatial unit. All those variables are
constant along the temporal period analized in this research. In this
subsetion we present some evidences of the expected relationship between
the four control variables considered an the incidence ratio.

The structure of population is a key variable on the incidence. Several
papers related the relationship between a fast trasnmision of COVID-19
and population density (e.g.~cumulative incidence in urban areas like
NYC) and therefore a positive sign will be expected. Note that our
analysis is at provincial level - not a urban level - and therefore is
necessary to rethinking the relationship. In Spain the population is
concentrate (with som exception) in the mediterranean coast (Goerlich
and Cantarino (2013)), in general with high temperature and high
humidity. So, positive or negative sign will be expected in the model.

The percentaje of older adults (over 65) is the second variable related
with the population structure include like control variable. Unlike the
population density the older people in Spain is concentrate in the core
of the Iberian Peninsula (see figure \ref{plotfer1}).

The per capita GDP also can generate differences between provinces. No
direct relationship can be observed between the GDP and the ratio of
incidence but the GDP is a proxy of inobservable information. This
varaible has been included in similar papers (Wang et al. (2020)). For
low DGPpc is more difficult maintain social distance (sing +); High
DGPpc -\textgreater{} more economic activity -\textgreater{} es mas
difícl mantener el confinamiento.

Posibles relaciones causales Alto GDP -\textgreater{} mas movimiento de
la población. Se recibió más gente de otros sitios que podría venir con
el virus puesto! Alto GDP -\textgreater{} La actividad económica de la
industria y la construcción solo se paralizó unos días. Aquellas
provincias con alto GDP tienen mas astividad industrial/contrucción
-\textgreater{} más movimimiento de la población. Bajo GDP
-\textgreater{} La población con menores ingresos no puede ser tan
obediente a la orden de confinamiento (suele ser menos disciplinada).

@@

@@

@@

@@

\leavevmode\hypertarget{ref-wang2020high}{}%
Wang, J., Tang, K., Feng, K., Lv, W., 2020. High temperature and high
humidity reduce the transmission of covid-19. Available at SSRN 3551767.

@@

@@

>>>>

  \begin{abstract}
  Spreading with astonishing speed, the novel SARS-CoV2 has swept the
  globe, causing enormous stress to health systems and prompting social
  distance guidelines and mandates to arrest its progress. While there is
  encouraging evidence that early public health interventions have slowed
  the spread of the virus, this has come at a high cost as the global
  economy is brought to its knees. How and when to ease restrictions to
  movement hinges in part on the question whether SARS-CoV2 will display
  seasonality associated with variations in temperature, humidity, and
  hours of sunshine. In this research, we address this question by means
  of a spatial analysis of the incidence of COVID-19 in the provinces in
  Spain. Use of a spatial Seemingly Unrelated Regressions (SUR) approach
  allows us to model the incidence of reported cases of the disease per
  100,000 population, as a function of temperature and humidity, while
  controlling for GDP per capita, population density, percentage of older
  adults in the population, and presence of mass transit systems. An
  interesting aspect of the spatial SUR approach is that it models
  incidence as a contagion process. Our results indicate that incidence of
  the disease is lower at higher temperatures and higher levels of
  humidity, although coefficients for this variable are significant only
  in some equations. Sunshine, in contrast, displays a positive
  association with incidence of the disease. Our control variables also
  yield interesting insights. Higher incidence is associated with higher
  GDP per capita and presence of mass transit systems in the province; in
  contrast, population density and percentage of older adults display
  negative associations with incidence of COVID-19.
  \end{abstract}

@@

Some of what is thought about the possible seasonality of COVID-19 is
based on analogies to the patterns of other known respiratory viruses.
However, de Ángel Solá et al.~(2020) note that ``not all seasonal
respiratory viruses experience the same spatiotemporal patterns''
(section 4). This urges caution when extrapolating from known viruses,
and indeed, the evidence in this respect is inconclusive. At a global
scale, de Ángel Solá et al.~(2020) see less risk in the Caribean Basin;
however, Coelho et al.~(Coelho et al., 2020) warn that at least during
the exponential phase, expansion of the virus is not driven by climate.
Similarly, whereas Araujo and Naimi (2020) argue that spread of
SARS-CoV2 will likely be constrained by climate, Harbert et al.~(2020)
remain unconvinced that spatial modelling can currently discriminate the
distribution of the disease on the basis of climate, at least in the
United States. Yao et al.~(2020), examined data from China and came to
the conclusion that neither temperature nor ultraviolet indices had an
association with transmission of COVID-19. This is despite previous
research that has linked less exposure to UVB radiation to higher
prevalence and severity of acute respiratory tract infections
(Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen et
al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
Moriyama, Hugentobler, and Iwasaki 2020). To further complicate matters,
much of the relevant work has yet to be peer-reviewed and therefore is
open to change (see for example the challenge of Harbert et al. (2020)
to Araujo and Naimi (2020)). In the United States, the National Academy
of Sciences, Engineering, and Medicine was engaged by the Office of the
Executive for guidance on this matter (see National Academies of
Sciences, Engineering and Medicine, 2020). Their conclusion summarizes
the situation well (see p.~6): ``Some limited data support a potential
waning of cases in warmer and more humid seasons, yet none are without
major limitations\ldots Additional studies as the SARS-CoV-2 pandemic
unfolds could shed more light on the effects of climate on
transmission.''

@@

https://github.com/paezha/covid19-environmental-correlates/tree/master/Environmental-Correlates-of-COVID19-Spain

@@

The global emergence of infectious diseases is mostly driven by
environmental, ecological, and socio-economic factors (Jones et al.,
2008). In the case of SARS-CoV2, the ecological factors include the
interaction between humans and wildlife. Once transmission of a disease
begins to happen between humans, socio-economic and environmental
factors become increasingly important. As noted in the introduction, the
focus of the paper is on environmental variables, namely temperature,
humidity, and sunshine. These variables have been implicated in the
viability and ease of transmission of similar viruses. In addition to
these variables, we also aim to use a set of controls, in the form of
specific socio-economic and demographic characteristics of each
province.

The first variable that we consider is GDP per capita. Much has been
said about globalization and the spread of infectious
disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p. A18)}.
The growth in global connections has presented a challenge to spatial
approaches in the initial stages of disease management, when its cause
may still be unclear (Zhou and Coleman, 2016). In reference to the
earlier outbreak of SARS, van Wagner (Van Wagner, 2008) remarks how
Toronto's status as a global city was a vulnerability. In our case, we
think of GDP per capita as a marker of a region's relative position in a
network of global cities, and their potential to be further ahead in the
trajectory of the pandemic. Furthermore, wealthier regions also tend to
concentrate more activities that produce non-traded goods, including
building and construction (Hallet, 2002). Therefore, it is possible that
wealthier regions remained relatively more active even during the
lockdown. On the other hand, it is also possible the less wealthy
regions have a higher proportion of workers in manual occupations that
cannot telework.

The percentaje of older adults (over 65) is the second variable related
with the population structure include like control variable. Unlike the
population density the older people in Spain is concentrate in the core
of the Iberian Peninsula (see figure \ref{plotfer1}). Our hypothesis is
that in the provinces with more older adults the social contact is lower
(Prem et al. (2017)) and therefore the a negative correlation will be
expected. This hypotesis is based in the idea that the agee-structured
social contact matrix in Spain is similar to Italy (see Singh and
Adhikari (2020) Fig 1 pag 3)

The structure of population is a key variable on the incidence. Several
papers related the relationship between a fast trasnmision of COVID-19
and population density (e.g.~cumulative incidence in urban areas like
NYC) and therefore a positive sign will be expected. Note that our
analysis is at provincial level - not a urban level - and therefore is
necessary to rethinking the relationship. In Spain the population is
concentrate (with some exception) in the mediterranean coast (Goerlich
and Cantarino (2013)), in general with high temperature and high
humidity. So, positive or negative sign will be expected in the model.

@@

Transit.

@@

\leavevmode\hypertarget{ref-Hallet2002regional}{}%
Hallet, M., 2002. Regional specialisation and concentration in the eu,
in: Regional Convergence in the European Union. Springer, pp. 53--76.

@@

\leavevmode\hypertarget{ref-Jones2008global}{}%
Jones, K.E., Patel, N.G., Levy, M.A., Storeygard, A., Balk, D.,
Gittleman, J.L., Daszak, P., 2008. Global trends in emerging infectious
diseases. Nature 451, 990--993.
doi:\href{https://doi.org/10.1038/nature06536}{10.1038/nature06536}

@@

\leavevmode\hypertarget{ref-singh2020age}{}%
Singh, R., Adhikari, R., 2020. Age-structured impact of social
distancing on the covid-19 epidemic in india. arXiv preprint
arXiv:2003.12055.

\leavevmode\hypertarget{ref-vanWagner2008toward}{}%
Van Wagner, E., 2008. Toward a dialectical understanding of networked
disease in the global city: Vulnerability, connectivity, topologies.
Networked disease: Emerging infections in the global city 13--26.

@@

\leavevmode\hypertarget{ref-Yao2020association}{}%
Yao, Y., Pan, J., Liu, Z., Meng, X., Wang, W., Kan, H., Wang, W., 2020.
No association of covid-19 transmission with temperature or uv radiation
in chinese cities. European Respiratory Journal 2000517.
doi:\href{https://doi.org/10.1183/13993003.00517-2020}{10.1183/13993003.00517-2020}

@@

\leavevmode\hypertarget{ref-Zhou2016accelerated}{}%
Zhou, Y.R., Coleman, W.D., 2016. Accelerated contagion and response:
Understanding the relationships among globalization, time, and disease.
Globalizations 13, 285--299.
